---
title: "REVISITING STUDY DESIGN"
subtitle: "The quest for validity"
author: "Austin Hart"
institute: "American University"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, rladies, rladies-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  message=FALSE, warning=FALSE, eval = TRUE, echo = FALSE, 
  fig.align = 'center', dev='svg'
)
```


```{r results='hide'}
library(tidyverse)
library(magrittr)
library(kableExtra)

setwd("C:\\Users\\ahart\\Dropbox\\SIS 600 AH\\2021S 600\\content\\w07 associations\\GroupDiffs")

load('DCPS testing.RData')
```

# Plan for today

Rethinking your approach to causal inference.  

- Internal validity and research design
  - Hurdles to causality
  - Designs for inference
  - Evaluating claims in practice

- Ex 1: Quota removal and representation

- Ex 2: Voting Rights Act


---
class: middle, right, inverse

# INTERNAL VALIDITY
### and how to find it


---
# The road to internal validity

.left-column[

### What is it?


]

.right-column[

> A causal claim is internally valid if it identifies the true, unbiased effect of a treatment, T, on an outcome, Y. 
> 
> Otherwise, the claim is biased or spurious. 

]

---
# The road to internal validity

.left-column[

### What is it?

### Criteria for assessment

]

.right-column[

- Evidence of **empirical association** between T and Y? 

- **Causal mechanism** linking exposure to T and change in Y 

- Rules out **reverse causal claim** that Y is the cause of T 

- Controls for/Conditions on **potential confounds**. 

Note: different criteria for ITS designs.

]


---
# The road to internal validity

.left-column[

### What is it?

### Criteria for assessment

### Role of design

]

.right-column[

- Experiment (RCT)
  - Randomly assigns cases to T
  - Control!! (over confounds)  

- Interrupted Time Series (ITS)
  - Response to non-random intervention
  - Beware unique "trend" issues
  - Comparison cases are critical

- Observational
  - Simply record T and Y
  - Then fight like hell
  

]



---
# The road to internal validity

.left-column[

### What is it?

### Criteria for assessment

### Role of design

### In practice

]

.right-column[

- Diagnose the causal claim 

- Identify the study design 

- Locate evidence of association 

- Work through the hurdles. 

]


---
class: inverse, middle, right

# Quotas and representation
### Bhavnani (2009)


---
# Diagnosing a claim (Bhavnani)

.left-column[

### Main question

]

--

.right-column[
  
- What is the effect of electoral reserves on women's representation after the reserve is withdrawn?

- More broadly, how does seeing/experiencing a woman in office influence attitudes about women in office?

]

---
# Diagnosing a claim (Bhavnani)

.left-column[

### Main question

### Design 

]

--

.right-column[
  
- Experiment: India randomly reserves 1/3 of local seats for women **at random** each election cycle

- Units: wards in Mumbai, India

- Key comparison
  - 81 *control* wards: no reserve in 1997 or 2002
  - 37 *treatment* wards: reserve in 1997 removed in 2002
]


---
# Diagnosing a claim (Bhavnani)

.left-column[

### Main question

### Design 

### Evidence

]

.right-column[
  
> Do "treated" wards (relative to control wards) see more women elected, more women running, etc *after* the reserve is gone?
  
  
See Table 3.
  
]



---
# Diagnosing a claim (Bhavnani)

.left-column[

### Main question

### Design 

### Evidence

### Validity

]

.right-column[
  
- Association?

- Mechanism?

- Time order?

- Confounding bias?

]


---
class: inverse, middle, right

# Literacy tests and the VRA
### Cascio and Washington (2014)


---
# Diagnosing a claim (Cascio and Washington)

.left-column[

### Main question

]

--

.right-column[
  
- What is the effect of expanding black voting rights on blacks' share of public spending?

- Specifically, how did the elimination of literacy tests impact state spending on education?

]

---
# Diagnosing a claim (Cascio and Washington)

.left-column[

### Main question

### Design 

]

--

.right-column[
  
- ITS: VRA eliminates literacy tests in Southern states that still used them at the time.  

- Units: counties in US Southern states

- Key comparison
  - 477 *control* counties: NO lit test in 1965
  - 625 *treatment* counties: YES literacy test
]


---
# Diagnosing a claim (Cascio and Washington)

.left-column[

### Main question

### Design 

### Evidence

]

.right-column[
  
> After the VRA, did state transfers to higher black population counties increase more in "treated"/literacy test counties (relative to control counties)?
  
  
See, e.g., Figures 1 and 2.
  
]



---
# Diagnosing a claim (Cascio and Washington)

.left-column[

### Main question

### Design 

### Evidence

### Validity

]

.right-column[
  
- History?

- Maturation?

- Regression?

- Instability?

- Instrumentation or testing?

]


---
class: inverse

# Wrapping up

### Internal validity

- It's a researcher thing; not a software thing

- Inextricably linked to research design

- No "1 grand test" of validity

